---
title: "ReproducibleResearch"
output: html_document
---

##Load packages
```{r packages, echo=FALSE}
library(tidyverse)
library(lubridate)
```

##Read in the Data files from the source
```{r data}
Temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",Temp)
data <- read_delim(unzip(Temp),",",col_types = list(
  steps = col_double(),
  date = col_date(),
  interval = col_double()
)) 
unlink(Temp)
```

## Problem 1: What is mean total number of steps taken per day?
### Problem 1 - Step 1: 
Calculate the total number of steps taken per day
```{r p1s1}
data.p1s1 <- data %>%
  group_by(date) %>%
  summarise(TotalSteps = sum(steps,na.rm = T), .groups = "drop") 
```

### Problem 1 - Step 2: 
Make a histogram of the total number of steps taken each day
```{r p1s2, echo= FALSE}
ggplot(data = data.p1s1, mapping = aes(x =TotalSteps)) +
  geom_histogram(color = "blue",binwidth = 2000) +
  ggtitle( "Steps Per Day")
```

### Problem 1 - Step 3: 
Calculate and report the mean and median of the total number of steps taken per day
```{r p1s3, echo=FALSE}
data.p1s2 <- data.frame( DailyMean = mean(data.p1s1$TotalSteps,na.rm = T), 
                         DailyMedian = median(data.p1s1$TotalSteps, na.rm = T))
print(data.p1s2)
```

## Problem 2: What is the average daily activity pattern?
### Problem 2 - Step 1: 
Determine the Average steps for each interval
```{r p2s11}
data.p2s1 <- data %>%
  group_by(interval) %>%
  summarise(Ave_Interval_Steps = mean(steps,na.rm = T), .groups = "drop")
```
Make a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r p2s12, echo=FALSE}

with(data.p2s1,plot(interval, Ave_Interval_Steps,type = "l"
               ,ylab = "Average Steps Per 5 Min Interval"
               ,xlab = "Military Time"
               ,main = "Ave Steps/per 5 min Interval"
               ,col = "blue"))
```

### Problem 2 - Step 2: 
Which 5-minute interval, on average across all the days in the dataset, contains the 
maximum number of steps?
```{r p2s2}
data.p2s1[which.max(data.p2s1$Ave_Interval_Steps),]
```

## Problem 3: Imputing missing values
### Problem 3 - Step 1: 
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs
```{r p3s1}
sapply(data,function(x){sum(is.na(x))})
```

### Problem 3 Step 2:
Replace the NAs with the average steps for the interval/day

### Problem 3 Step 3: 
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r p3s3}
data.p3s2 <- data %>%
  mutate(day = wday(date)) %>%
  group_by(interval,day) %>%
  mutate(steps = ifelse(is.na(steps),mean(steps,na.rm = T),steps)) %>%
  ungroup() %>%
  select(-day)
```

### Problem 3 Step 4: 
Make a histogram of the total number of steps taken each day with and without NA values
```{r p3s41, echo=FALSE}
data.p3s4 <- data.p3s2 %>%
  group_by(date) %>%
  summarise(TotalSteps = sum(steps), .groups = "drop")

par(mfrow = c(1,2))
with(data = data.p3s4, hist(TotalSteps
     , main = "NAs replaced"))
with(data = data.p1s1, hist(TotalSteps
     , main = "NAs excluded"))
```

Notice the mean and median total number of steps taken per day increase as a result of imputing missing data on the estimates of the total daily number of steps.
```{r p3s42}
data.p3s4.sum <- data.frame( DailyMean = mean(data.p3s4$TotalSteps), 
                         DailyMedian = median(data.p3s4$TotalSteps))
data.p3s4.sum
data.p1s2
```
##Problem 4: Are there differences in activity patterns between weekdays and weekends?
###Problem 4 - Step 1: 
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r p4s1}
data.p4s1 <- data.p3s2 %>%
  mutate(weekend = factor(
    ifelse(weekdays(date) 
           %in% c("Saturday","Sunday")
           ,"Weekend"
           ,"Weekday")))
```

##Problem 4 - Step 2: 
Make a panel plot containing a time series plot (i.e.type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r p4s2}
data.p4s2 <-data.p4s1 %>%
  group_by(weekend) %>%
  mutate(TotalAve = mean(steps))%>%
  group_by(interval, weekend,TotalAve) %>%
  summarise(AveSteps = mean(steps), .groups = "drop")

ggplot(data = data.p4s2, mapping = aes(x = interval, y = AveSteps, color = weekend)) +
  geom_line() +
  geom_hline(mapping = aes(yintercept = TotalAve, color = weekend))+
  facet_wrap(~weekend,nrow = 2)
```
